---
title: "Logistic Regression Foundations"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
  messages: true
---

In the general linear model framework (either simple or multiple linear
regression), the goal is to model the mean response or outcome given a set of
predictors. This is often written as 
$$
\hat{y} = \beta_0 + \beta_1x_1 + \cdots + \beta_px_p,
$$
where $\hat{y}$ is the estimated mean response. The goal of the linear model is
to produce estimates for the parameters $\beta_i$ that produce the line of best
fit to the data. The equation of the line of best fit represents the predicted
mean response, $\hat{y}$ given the predictors.

But what if the outcome variable is not continuous, but is rather a categorical
variable? Perhaps the outcome is whether a patient has some disease or
if a pest insect is present in an agriculture plot. In these cases the outcome
could be represented as 0/1, Yes/No, True/False or some other binary measure.

## The Dataset
The dataset that will be used in this section is the `chdage` date from the
`aplore3` package that supports the text Applied Logistic Regression by Hosmer,
Lemeshow and Sturdivant. (2013) It contains data on coronary heart disease (CHD)
versus age for a high risk population. We will fit a very simple logistic
regression model to this data to explore the very basics of how logistic
regression works.

Before fitting any models, an exploratory data analysis should be conducted on
the data. This serves a few purposes. First, it is form of quality assurance. 
Are there obvious problems with the data like miscoded values or large amounts
of missing data? Is the data in a format that can be used directly by the type
of models we might want to fit to the data? How are the variables distributed?

For now, we will just make a couple of checks starting by a quick glimpse of the
data.
```{r initial setup}
rm(list=ls())
options(scipen=999) # turn off scientific notation
library(tidyverse); library(aplore3); library(latex2exp)
theme_set(theme_bw()) # use the black & white theme for all plots
# glimpse the data
chd <- aplore3::chdage
glimpse(chd)
```
Right away, we can see that there are 4 variables. The `id` variable won't be of
much use to us. We can see that there are two factor variables: `chd` and
`agegrp`. We are going to want to convert the `chd` variable to numeric with
values 0 and 1 so that we can plot the values numerically as well as fit a
linear model, for demonstration purposes.

The `glimpse()` function also shows us that the `agegrp` factor has an issue
with its levels. There is a level `20-39` that overlaps with the level `30-34`.
A quick look at the levels, `r levels(chd$agegrp)` shows that the level `20-39`
should be `20-29`, so this should be fixed. We won't actually use this variable,
but we'll make the correction anyway.

We will create two new variables, `CHD` and `AgeCat`, that address these issues.

```{r data prep}
# prep/clean-up the data for use
chd <- chd |> mutate(
    # for plotting, make the response a numeric value rather than categorical
    CHD = if_else(chd == "Yes", 1, 0),
    # there is an error in the data set where the 20-29 level is indicated as 20-39
    AgeCat = if_else(agegrp == "20-39", "20-29", agegrp) |> as_factor()
  ) |> select(CHD, "Age"=age, AgeCat)
```

Now, the data is ready to use so let's take a look and make a first attempt
at a model fit.

## Plotting the Data and a Linear Model
Consider the case where the outcome indicates whether a patient has coronary 
heart disease or not. If the outcome is coded as 0=no disease and 1=disease, a
plot of disease presence vs age could look like this:

```{r plot presence v age}
#| fig-align: "center"
#| fig-cap: "Presence/absence of coronary heart disease in $n=100$ patients aged 20 to 70 year."
#| fig-height: 2.5
ggplot(chd, aes(x=Age, y=CHD)) + geom_point()
```
We could fit a linear model to this data to try to estimate the mean probability
of having CHD at each age. The `lm()` function will happily estimate the
parameters $\beta_0$ and $\beta_1$ for us, and we can plot the line of best fit
(in purple) to the data.

$$
\hat{\pi}_i = \beta_0 + \beta_1\cdot\text{Age}_i.
$$

:::{.callout-note}
Probability will be denoted by the Greek lettre $\pi$. You often see $p$ used,
but $p$ can represent a p-value or the number of predictors in a model ($x_p$,
for example), so we'll use $\pi$ instead.
:::

```{r linear fit}
#| echo: false
#| fig-align: "center"
#| fig-height: 2.5

lm.fit <- lm(CHD~Age, data=chd)
m <- coef(lm.fit)[2]; b <- coef(lm.fit)[1]
eqn <- paste("$\\widehat{\\pi}_{CHD,i}=", round(b,2)," + ", round(m,2), "Age_i$")
ggplot(chd, aes(x=Age, y=CHD)) + geom_point() + 
  geom_abline(slope=m, intercept=b, colour="orchid4", linewidth=1) +
  annotate("text", x=37,y=0.5, label=TeX(eqn, italic=TRUE), colour="orchid4")
```
:::{.callout-note}
Can you see the problems with this approach?
:::

There are a couple of clues that trying model probabilities with a linear model
will not work well. First, on the low end of the age range, the model predicts
negative probabilities for the presence of CHD. Additionally, it predicts
probabilities greater than 1 for ages above about 71 years. These results
clearly don't make any sense.

What we would really like is for probabilities to be bound between 0 and 1 and
for the line/curve of best fit to asymptotically approach these boundary values.
A logistic curve can satisfy these constraints. Here is what a generic logistic
curve looks like:
```{r}
#| echo: false
#| fig-align: "center"
#| fig-height: 2.5
x <- 0:100
y <- exp(-5  + 0.1*x)/(1+exp(-5  + 0.1*x))
ggplot(tibble(x,y), aes(x,y)) + geom_line(colour="orchid4", linewidth=1) +
  labs(y=TeX("$f(x)$"))+
  theme(
    axis.text=element_blank(),
    axis.ticks=element_blank()
  )
```
Notice that this function has a bounded range and approaches the boundaries 
without ever reaching them.

## The Logistic Model
The logistic function looks like a good candidate for our scenario, but how is
it described mathematically? First, some definitions. Odds play a prominent role
in logistic regression, but how are they calculated? It's actually quite easy.
Odds are defined as the ratio of the probability of an event occurring to the
probability of that event not occurring. Unlike probabilities which can only
range between 0 and 1, the odds of an event can range from 0 to infinity. If the
probability of an event occurring is denoted $\pi$, then the odds are defined as

$$
\text{odds}=\frac{\pi}{1-\pi}.\tag{1}
$$
If we rearrange Equation 1, we get an expression for the probability of the 
event, $\pi$ in terms of the odds,
$$
\pi = \frac{\text{odds}}{1 + \text{odds}}.\tag{2}
$$
Now that odds and probability have been defined in terms of each other, we can
think a bit about how we might model probabilities with the logistic function.
But before we can do that, we need to know the functional form of the logistic
function and how we can relate it to odds and probabilities.

The logistic function has this basic form,
$$
f(x) = \frac{e^x}{1+e^x}.
$$
The $e^x$ term has range $(0,\infty)$, which is a property we know is desirable
for us: it matches the behaviour of odds. If we think about the denominator, we
can that as $e^x$ gets really large, then the numerator is *almost* as big as
the denominator, but not quite. Thus, as $x\to\infty$, the logistic function
gets closer and closer to 1, but can never equal or exceed 1. Similarly, as
$x\to -\infty$, then $e^x\to 0$, which means that the logistic function is
bounded below by 0. This is good news. The logistic function can take on the
same range of values as probabilities, which is exactly what we're looking for.

Notice that the form of the logistic function looks very similar to the
definition of probability in terms of odds given in Equation 2. If we
let $\text{odds}=e^x$, then
$$
\pi = \frac{e^x}{1+e^x}.
$$
This is a good, start, but what we really want is the conditional probability of the outcome, $Y$, given 
some set of predictors. If we have collected data on the
outcome variable and $p$ predictors, then we could replace the placeholder $x$
in the equation above with the expression 
$\beta_0 + \beta_1x_1 + \cdots + \beta_px_p$. Now, the probability equation is
$$
\pi_i = \frac{e^{\beta_0 + \beta_1x_{1,i} + \cdots + \beta_px_{p,i}}}{1+e^{\beta_0 + 
\beta_1x_{1,i} + \cdots + \beta_px_{p,i}}}.\tag{3}
$$
Equation 3 is known as the logistic response function. You may also see it 
written as 

$$
\pi_i = \frac{1}{1+e^{-(\beta_0 +
\beta_1x_{1,i} + \cdots + \beta_px_{p,i})}}.
$$

:::{.callout-note}
Can you show that these two formulations of the logistic response function are
equivalent?
:::

At the moment, we have a non-linear function. It would be even better if we could
model the probability of the outcome given the predictors as a linear function.
Recall our definition of odds from above. If we look at Equation 3, we see the
odds are given by $e^{\beta_0 + \beta_1x_{1,i} + \cdots + \beta_px_{p,i}}$. But
using our definition of the odds, we can construct the following equation:
$$
\frac{\pi_i}{1-\pi_i}=e^{\beta_0 + \beta_1x_{1,i} + \cdots + \beta_px_{p,i}}.\tag{4}
$$
We can show that Equations 3 and 4 are equivalent with some algebra.

$$
\begin{aligned}
\pi &= \frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}\\
\frac{\pi}{1-\pi} &= \frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}\cdot\frac{1}{1-\pi}\\
\frac{\pi}{1-\pi} &= \frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}\cdot\frac{1}{1-\frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}}\\
\frac{\pi}{1-\pi} &= \frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}\cdot\frac{1}{(\frac{1}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}})(1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}-e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p})}\\
\frac{\pi}{1-\pi} &= \frac{e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}\cdot\frac{1+e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}}{1}\\
\frac{\pi}{1-\pi} &= e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p}.
\end{aligned}
$$
If we take the log of both sides of the equation, 
$$
\log\bigg(\frac{\pi}{1-\pi}\bigg)=\log(e^{\beta_0 + \beta_1x_1 + \cdots + \beta_px_p})= \beta_0 + \beta_1x_1 + \cdots + \beta_px_p,
$$
we have a function of the log odds that is linear in terms of the
parameters $\beta_i$. The expression on the left side of the equation is often
called the logit and it is common to see the above equation written as
$$
\text{logit}(\pi)=\beta_0 + \beta_1x_1 + \cdots + \beta_px_p.\tag{5}
$$

## Fitting a Simple Logistic Regression Model
Returning the coronary heart dataset, suppose that we want to fit a model that
predicts the presence of CHD in a high risk population of adults using their age
as the explanatory variable. The formal model will be
$$
\begin{aligned}
Y_i&\sim bin(1, \pi_i)\\[0.25cm]
\pi_i &= \frac{\exp(\eta_i)}{1+\exp(\eta_i)}=\frac{1}{1 + \exp(-\eta_i)}\\[0.25cm]
\eta_i &= \beta_0 + \beta_1\cdot Age_i.
\end{aligned}
$$

The model fit is carried out with the `glm()` function in R. The arguments to this
function are similar to those of the `lm()` function for fitting linear models
with a normal response, with the exception of the `family` argument. This
argument tells the `glm()` function that the response has a binomial
distribution and that the `link` function is the logit. The link function
relates the mean response to the linear equation of parameters. In the standard
linear model, the link function is the identity function, so the mean is
directly related to the linear function of parameters. In logistic regression,
the mean response ($\pi$) is related to the linear predictor ($\eta$) by the
logit function. In other words, $\text{logit}(\pi_i)=\eta_i$. The expression
`family=binomial(link=logit)` in the `glm()` call below is R's way of telling
`glm()` that we want to fit a logistic regression model described by the formula
`CHD ~ Age` with the dataset `chd`.

Running the R code (click Show the code to see the actual R code used to fit the model)
below produces the following model output:

```{r fitting a LR}
lr.fit <- glm(CHD ~ Age, data=chd, family=binomial(link=logit))
summary(lr.fit)
```

How do we interpret this model summary?

### Interpreting the Estimated Model Parameters
For now, we're going to limit ourselves to the `coefficients` section of the
model summary output and in particular the `Estimate` column. The estimated model equation is $\text{logit}(\hat{\pi}_i) = -5.31 + 0.11\cdot\text{Age}_i$. 
The intercept term is $\hat{\beta}_0=-5.31$ and the slope coefficient associated
with age is $\hat{\beta}_1=0.11$. The $\hat{ }$ symbol over the two $\beta$
terms indicates that the value is an estimate of the unknown population
coefficient $\beta_i$.

The intercept parameter estimate tells us the log odds of having CHD when `Age`
is zero years:

$$
\text{logit}(\hat{\pi}_i) = \hat{\beta}_0 + \hat{\beta}_1\cdot\text{Age}_i=-5.31 + 0.11(0) = -5.31.
$$
If we exponentiate both sides of this equation, we obtain the odds of having CHD when aged 0 years:
$$
\exp[\text{logit}(\hat{\pi}_i)] = \exp[\hat{\beta}_0 + \hat{\beta}_1\cdot\text{Age}_i]=\exp[-5.31 + 0.11(0)] = \exp[-5.31] = 0.00494.
$$
The intercept is not meaningful in this case as an age of zero years doesn't 
make sense. We could change this by centring the age variable around its mean. 
Using a centred age, the intercept would present an individual who is of the average
age in the dataset and this *would* be meaningful. Centring `Age` will not change
the value of the slope parameter estimate, $\hat{\beta}_1$. We will look at 
centring data in a more realistic example soon.

How do we interpret the slope parameter estimate for `Age`, $\hat{\beta}_1$? We
see that the log odds of having CHD increase by 0.11 for every one year increase
in age. Consider the log odds of having CHD for an individual of age $\text{Age}_i + 1$
compared to an individual of age $\text{Age}_i$.
$$
\begin{aligned}
\log\bigg(\frac{\hat{\pi}_i}{1-\hat{\pi}_i}\bigg) &= \hat{\beta}_0 + \hat{\beta}_1\cdot(\text{Age}_i + 1)\\[0.25cm]
\log\bigg(\frac{\hat{\pi}_i}{1-\hat{\pi}_i}\bigg) &= \hat{\beta}_0 + \hat{\beta}_1\cdot(\text{Age}_i)
\end{aligned}
$$
We can compare these two log odds with a ratio. To simplify the notation, the
odds for the individual of age $\text{Age} + 1$ will be denoted $O_{A}$ and the
odds for the individual of age $\text{Age}$ will be denoted $O_{B}$.

$$
\begin{aligned}
\log\bigg(\frac{O_A}{O_B}\bigg) &= \log(O_A) - \log(O_B)\\[0.25cm] 
&= \hat{\beta}_0 + \hat{\beta}_1\cdot(\text{Age}_i + 1) - [\hat{\beta}_0 + \hat{\beta}_1\cdot(\text{Age}_i)]\\[0.25cm]
&= \hat{\beta}_1\cdot(\text{Age}_i + 1) - \hat{\beta}_1\cdot(\text{Age}_i)\\[0.25cm]
&= \hat{\beta}_1(\text{Age}_i + 1 - \text{Age}_i)\\[0.25cm]
&= \hat{\beta}_1
\end{aligned}
$$
The problem with log odds is that it's not very easy to interpret. Typically,
odds ratios are used instead because they are easier for most people to
understand. In the case of our model, if we want the odds ratio of developing
CHD for a one year increase in `Age`, we just exponentiate both sides of this
equation:

$$
\begin{aligned}
&\exp\bigg[\log\bigg(\frac{O_A}{O_B}\bigg)\bigg] =\exp[\hat{\beta}_1]\\[0.25cm]
&\frac{O_A}{O_B} = e^{\hat{\beta}_1} = e^{0.11} = 1.116.
\end{aligned}
$$
There are a couple of ways that this can be communicated to an audience. We
could say that the odds of CHD increase 1.16 times for a one year increase in
`Age`. We could also state the change as a percentage. The odds of the presence
of CHD increase 11.6% for a one year increase in `Age`.

To compute the percentage, simply subtract 1 from the odds ratio: 
$e^{\beta_1} - 1 = 0.116$ then multiply by 100 to get the percentage. If 
$e^{\beta_1} < 1$, then $e^{\beta_1} - 1 < 0$ and the odds of the presence of 
CHD would decrease with increasing `Age`.

## Using the Model to Make Predictions
We can get the estimated mean value, $\pi_i$ for our model using the
`predict()` function in R. Using `type=response` returns the probabilities. If we
wanted to get classifications (disease/no disease), we could specify 
`type=class` instead. More on classifications later. We can also compute the
estimated mean values using the logistic response function that was defined in
Equation 3. Comparing the output from the `predict()` function and the "hand"
calculation, the results are identical.

Let's say that we wanted to compute the predicted probability of having CHD for a
57 year old in the high risk population the observed data were sampled from. To
do this we can use the `predict()` function with a new vector of data.

Using R, we get the following result.
```{r prediction with R}
xh <- data.frame(Age=57)
predict(lr.fit, newdata=xh, type="response") |> unname()
```
We can check this against the simplified version of Equation 3:

$$
\hat{\pi}_h = \frac{1}{1+\exp[-(-0.531 + 0.11(57))]}
$$
```{r prediciton by hand}
# extract the parameter coefficient estimates from the model
b <- coef(lr.fit) |> unname()
1/(1 + exp(-(b[1] + b[2]*57)))
```

... and we get the same result "by hand". So, on average, a 57 year old in this
high risk population has about a 73% chance of having coronary heart disease.

Up next: We will look at a more complex model using the 
[Framingham Heart Study](multiple-logistic-regression.html) dataset.