---
title: "Mediators in Effects Models"
format: 
  html:
    code-fold: true
    code-summary: "Show the code"
bibliography: confounding-references.bib
execute:
  warning: false
  messages: false
---

```{r initial setup}
#| echo: false
rm(list=ls()); library(tidyverse)
options(scipen=999)
```

## What is a Mediator?
A mediator is the mechanism by which some feature of the world [^FoTW], the 
treatment, affects another feature of the world, the outcome. The effect of the
treatment may be thought of as "flowing through" the mediator in whole or in
part. The diagram below depicts this graphically.

![Path Diagram of a confounding reationship](images/mediation.svg){width=50%}
When constructing an effects model to investigate the effect of the treatment on
the outcome, it may be useful to include mediators in the model if a more 
granular analysis of the treatment effect is desired. However, if the total
effect of the treatment is the goal, then any mediators should be excluded from
the model. When constructing an effects model, it is important to understand the
relationships between the treatment variable and the other predictors in the
model be they mediators, [confounders](confounders.html) or 
[independent predictors](independent-predictors.html). The decision to include 
or exclude specific predictors should be weighed against their conceptual
relationships with the treatment and the outcome. Including predictors into the
model without forethought risks creating a biased estimate of the treatment
effect.

## Example - The Effect of Cylinder Count on Fuel Consumption
Imagine that you have been asked by an engine design company to investigate the
effect of the number of cylinders in an internal combustion engine and the fuel
consumption measured in miles per gallon (`mpg`). The company provides you a
dataset containing 6 variables on 32 different vehicle models. They have asked
you to build an effects model and to identify any important mediators and/or
independent predictors that help establish the effect of the number of engine
cylinders on fuel consumption. If there are any mediators, they would like to 
know the partial effect of these predictors as well.

The dataset you are provided has the following variables: `mpg`, `cyl`, `disp`, 
`wt`, `gear` and `am` where

- `mpg`: Miles/(US) gallon  
- `cyl`: Number of cylinders  
- `disp`: Displacement (cu.in.)  
- `wt`: Weight (1000 lbs)  
- `gear`: Number of forward gears  
- `am`: Transmission (0 = automatic, 1 = manual)  







```{r}
df <- mtcars |> 
  select(mpg, cyl, disp, wt, gear, am) |> 
  mutate(
    cyl = as_factor(cyl),
    gear = as_factor(gear),
    am = as_factor(am)
  )
lm(mpg~cyl+gear, data=df) |> summary()
lm(mpg~cyl+am, data=df) |> summary()
lm(mpg~cyl, data=df) |> summary()
```


```{r}
lm(mpg~cyl, data=df) |> summary()

lm(mpg~cyl+disp, data=df) |> summary()

lm(mpg~cyl+wt, data=df) |> summary()

lm(mpg~cyl+wt+disp+gear+am, data=df) |> summary()
```


## In Summary

Summarise what it means to be a mediator and the reasons that we might want to
include or exclude mediators from an effects model.

## References

::: {#refs}
:::

[^FoTW]: This expression comes from the excellent book Thinking Clearly with
Data. I find that it is a very intuitive and accessible way to refer to what
will become the columns of a dataset: the variables that we collect data on and
have research questions about. I will use it interchangeable with "variable"
throughout this guide. [@bueno_de_mesquita_thinking_2021]
