---
title: "Estimating the Effect of Smoking on Coronary Heart Disease"
format: 
  html:
    fig-height: 2.5
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
  messages: true
---

## The Dataset


```{r initial setup}
library(tidyverse)
theme_set(theme_bw())
chd <- read_csv("framingham.csv") |> 
  select(male, age, currentSmoker, totChol, diaBP, BMI, glucose, TenYearCHD)

chd <- chd[complete.cases(chd),]

chd <- chd |> 
  mutate(
    sex = if_else(male == 1, "Male", "Female") |> factor(levels=c("Female", "Male")),
    smoker = if_else(currentSmoker == 1, "Yes", "No") |> as_factor(),
    haschd = if_else(TenYearCHD == 1, "Yes", "No") |> as_factor()
  ) |> 
  select(-c(male, currentSmoker, TenYearCHD))

library(psych); library(gt)
# helper functions
univariate_summary <- function(data) {
  describe(data, IQR=TRUE) |> 
    select("Mean"=mean, "SD"=sd, "Median"=median, "Min"=min, "Max"=max, IQR)
}

univariate_plot <- function(data, field, xlab, bins=30) {
  ggplot(data, aes(x={{field}})) + 
    geom_histogram(aes(y = after_stat(density)), colour="black", bins=bins, fill="white", alpha=0.7) +
    geom_density(linewidth=1, fill="seagreen", alpha=0.2) + 
    labs(x=xlab)
}

```

```{r skim the dataset}
library(skimr)
skim(chd)
```


**Cite these guys:** [Framingham Heart Study - CHD Risk Prediction](https://github.com/KomlaAkpalu/Framingham-CoronaryHeartDisease-Modeling/tree/main)

## Univariate Analyses

```{r}
var_summaries <- bind_rows(
  univariate_summary(chd$age),
  univariate_summary(chd$totChol),
  univariate_summary(chd$diaBP),
  univariate_summary(chd$BMI),
  univariate_summary(chd$glucose)
)

rownames(var_summaries) <- 
  c("Age", "Total Cholesterol", "Diastolic BP", "Body Mass Index", "Blood Glucose")

var_summaries |> 
  gt(rownames_to_stub=TRUE) |> 
  fmt_number(decimals=1)
```


```{r}
#| fig-height: 6
#| fig-width: 10
library(patchwork); library(latex2exp)
(
  univariate_plot(chd, age, "Age (yr)", bins=10) + 
  univariate_plot(chd, totChol, "Total Cholesterol (mg/dL)") +
  univariate_plot(chd, diaBP, "Diastolic Blood Pressure (mm Hg)", bins=25)
)/(
  univariate_plot(chd, BMI, TeX("Body Mass Index ($kg/m^2$)")) +
  (univariate_plot(chd, glucose, "Blood Glucose (mg/dL)") +
  geom_vline(xintercept=126, linewidth=1, linetype="dashed", colour="grey30") +
  annotate("text", label="126 mg/dL cutoff for Diabetes diagnosis", x=140, 
           y=0.025, hjust=0.05))
) + plot_layout(axes = "collect")
```




## Bivariate Exploratory Analyses

```{r pairs plot}
library(GGally)
chd |> select(age, totChol, diaBP, BMI, glucose) |> ggpairs(progress=FALSE)
```





```{r}
boxplot(age ~ smoker, data=chd)
```




```{r}
# chisq.test(x=chd$male, y=chd$currentSmoker)
# table(chd$sex, chd$smoker) |> prop.table(margin=1)
```
Fitting an intial model with potentially important variables. It includes
variables that might be confounders, mediators and independent predictors.

```{r}
lr.modA <- glm(haschd ~ smoker + sex + age + totChol + glucose + BMI, 
    data=chd, family=binomial)
summary(lr.modA)
```
The first covariate we'll consider is `BMI`. Thinking about the potential causal
direction, it does not make sense that a person's BMI would influence whether
they smoke or not. However, smoking does influence `BMI` and thus `BMI` may
mediate smoking's influence on CHD risk.

```{r}
lr.modB <- glm(haschd ~ smoker + sex + age + totChol + glucose, 
    data=chd, family=binomial)
summary(lr.modB)
```
The removed of `BMI` did not substantially change the parameter coefficient
associated with smoking or its standard error. It is worth comparing the two
models to see if the model that includes BMI is preferred over the one that does
not.

```{r}
anova(lr.modB, lr.modA, test="LRT")
```
There is no evidence that the model that contains `BMI` is a better fit to the
data than the reduced model ($\chi^2_1=2.0788$, $p=0.1494$). Model B will become
our working model.

The next variable under consideration will be prevalence of hypertension at
baseline. Again, this variable is a likely mediator as the direction of
causality is from smoking to hypertension, not the other way around.

```{r model c}
lr.modC <- glm(haschd ~ smoker + sex + age + totChol + glucose, 
    data=chd, family=binomial)
summary(lr.modC)
```

```{r}
anova(lr.modC, lr.modB, test="LRT")
```
The likelihood ratio test indicates that `prevalentHyp` should be retained in
the model ($\chi^2_1=45.564$, $p<0.0001$). There does not appear to be a strong
mediating effect, the log odds of CHD actually decrease from 0.373 to 0.337
(about a 10% decrease) when `prevalentHyp` is removed from the model. While 
`prevalentHyp` appears to be potential predictor for predicting CHD, it doesn't
appear to be a confounder or a mediator. Since its inclusion in the model does
not appear to substantially change the standard error for `smoker`, it will be
removed from the model for reasons of parsimony.

Next, `glucose` is investigated as an independent predictor. There doesn't seem
to be any compelling association between blood sugar levels and smoking.

```{r model d}
lr.modD <- glm(haschd ~ smoker + sex + age + totChol, 
    data=chd, family=binomial)
summary(lr.modD)
```
Removal of glucose does not result in a major change to the estimated parameter
coefficient for smoking, nor does the standard error of the coefficient change
much.

```{r C vs D}
anova(lr.modD, lr.modC, test="LRT")
```
The likelihood ratio test suggests that `glucose` should be retained in the
model, but for estimating the effect of smoking on CHD, it does not seem
necessary to retain it.

There is some evidence that smoking increases total cholesterol and thus 
`totChol` would act as a mediator.


```{r model e}
lr.modE <- glm(haschd ~ smoker + sex + age, 
    data=chd, family=binomial)
summary(lr.modE)
```
There is very little change in the estimated parameter coefficient for smoking
when `totChol` is removed from the model. Standard errors change only very slightly.

```{r E vs D}
anova(lr.modE, lr.modD, test="LRT")
```
The likelihood ratio test indicates that `totChol` should be retained in the
model ($\chi^2_1=9.9658$, $p=0.0016$). Since there does not seem to be any real
mediating or confounding effect, it will be dropped for parsimony.

Next, the `age` variable will be considered. This variable along with `sex` is a
common confounder in many situations. We saw earlier that there are is a
systematic difference in the age of smokers and non-smokers, with non-smokers
tending to be younger than smokers. This leads me to suspect that there will be
a change in the apparent effect of smoking on the odds of having CHD if age is
removed from the model.

```{r model F}
lr.modE <- glm(haschd ~ smoker + sex, 
    data=chd, family=binomial)
summary(lr.modE)
```
There is a major change in the estimated parameter coefficient for smoking when
`age` is removed from the model. The effect of smoking on the odds of having CHD
essentially disappears. This is likely due to the fact that 

```{r}
ggplot(data=chd, aes(x=haschd, y=age, colour=smoker)) + geom_boxplot()
```


## Low Birth Weight - Smoking Effect

Things to consider when examining different proposed effects models for
quantifying the effect of smoking on the odds of a low birth weight:

- Conceptually, does it make sense that there would be an association between the treatment and the variable in question?
  - Is the variable in question a confounder or a mediator?
- Is the variable in question likely to be a strong independent predictor of low birth weight?

The first consideration relates to confounding. We want to keep confounders in
the model regardless of whether they improve the model or not. The goal is not
prediction here, it's understanding the effect of our treatment on the outcome.
Strong independent predictors may shrink the standard error of treatment, making
the estimate more precise. We definitely want to keep such predictors in the
model even though we aren't too interested in their role in predicting the
outcome because our focus is not prediction here.

### The Dataset

```{r}
library(aplore3)
glimpse(lowbwt)
```

```{r}
skimr::skim(lowbwt)
```


### Univariate EDA

```{r}
with(lowbwt, hist(age))
```

```{r}
with(lowbwt, hist(lwt))
```

```{r}
with(lowbwt, table(race))
```

```{r}
with(lowbwt, table(smoke))
```

```{r}
with(lowbwt, table(ftv))
```

```{r}
with(lowbwt, table(ptl))
```

```{r}
with(lowbwt, table(ht))
```


```{r}
with(lowbwt, table(ui))
```

